---
title: "H2Ohio_Connectivity_Analysis"
author: "Kenny Anderson"
date: "2026-07-10"
output: html_document
---
```{r, setup, include=FALSE}
options(scipen =999)
```

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(ggpubr)
library(matrixStats)
library(scales)
library(patchwork)
library(scales)
library(Metrics)
library(purrr)
library(stringr)
library(ggeffects)
library(ggspatial)  # for map annotations (north arrow, scale bar)
library(tidyterra)  # for handling and plotting raster data with ggplot2 (e.g., elevation)
library(pals)       # for color palettes (kelly)
library(sf)         # required to handle spatial data for maps
library(maps)       # for map data (e.g., county boundaries)
library(maptiles)   # for basemaps (e.g., Stamen Terrain)
library(ggrepel)    # for improved label placement in maps
library(magick)     # for image manipulation (e.g., adding logos to maps)
library(tigris)     # for Ohio boundary
```

## Load in Data

```{r}
annual_df = read.csv("Data/H2Ohio_2025_annualdf.csv")
month_df = read.csv("Data/H2Ohio_2025_monthdf.csv")
```


# Methods

### Wetland Summary Stats:
Get some summary stats on wetland hydrology
```{r}
loi_df = read.csv("Data/locations_of_interest_20260317.csv")
char_df = read.csv("Data/project character_20251119.csv")

# List all projects that have budgets
project_list <- unique(c(annual_df$Project, "MERW")) # add on MERW so we also capture values that don't list the three sub-projects individually
# subset out projects with budgets from char_df
char_df_budgets <- char_df %>%  filter(project_code %in% project_list) %>%  distinct(project_code, .keep_all = TRUE) # keep only the first row for each value

# We don't have EFLA and WALC in this db so we'll need to add values on manually. Also no data for UPPB so we'll need to add that too. 

# isolated pools present
char_df_budgets %>% summarise(n_yes = sum(isolated_pools == "Yes", na.rm = TRUE))
# number of lake connections
char_df_budgets %>% summarise(n_yes = sum(direct_coastal_connection == "Yes", na.rm = TRUE))
# number of stream connections
char_df_budgets <- char_df %>% filter(project_code %in% project_list) %>%  filter(str_detect(direct_riparian_connection, regex("Yes", ignore_case = TRUE)))
char_df_budgets %>% distinct(project_code) %>% pull(project_code)


# now do the same sorting with the LOI dataset to narrow in on the hydrology section
# subset out projects with budgets from loi_df that have direct tile inputs
loi_df_budgets <- loi_df %>%  filter(project_code %in% project_list) %>% filter(str_detect(hydrology, "direct tile input"))
# tell each project with direct tile inputs: 
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)
# subset out projects with budgets from loi_df that have floodplains/lotic influence
loi_df_budgets <- loi_df %>% filter(project_code %in% project_list) %>%  filter(str_detect(hydrology, regex("floodplain|lotic", ignore_case = TRUE)) | project_code == "EFLA") # manually include EFLA because it doesn't have data in the hydrology column
# tell each project with floodplains/lotic influence: 
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)


# subset out just pools
loi_df_budgets <- loi_df %>%  filter(project_code %in% project_list) %>% filter(str_detect(loi_cat_primary, regex("Pool|pool", ignore_case = TRUE)))
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)

```



# Results

### Nutrient Budget Summary Stats: 

```{r}
# count the number of annual budgets that are positive and negative
annual_df %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
annual_df %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

annual_df %>% group_by(Project) %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
annual_df %>% group_by(Project) %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

# count the number of months that are positive and negative
month_df %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
month_df %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

month_df %>% group_by(month) %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
month_df %>% group_by(month) %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

# mean and sd of annual retention: 
annual_df %>% summarise(mean_tp_lbs = mean(tp_filt_lbs), med_tp_lbs = median(tp_filt_lbs),sd_tp_lbs = sd(tp_filt_lbs), max_tp_lbs = max(tp_filt_lbs), min_tp_lbs = min(tp_filt_lbs))
annual_df %>% summarise(mean_tn_lbs = mean(tn_filt_lbs), med_tn_lbs = median(tn_filt_lbs), sd_tn_lbs = sd(tn_filt_lbs), max_tn_lbs = max(tn_filt_lbs), min_tn_lbs = min(tn_filt_lbs))

# mean and sd of nutrient uptake efficiency: 
annual_df %>% summarise(mean_tp_effic = mean(tp_effic,na.rm=T), med_tp_effic = median(tp_effic,na.rm=T), sd_tp_effic = sd(tp_effic), max_tp_effic = max(tp_effic), min_tp_effic = min(tp_effic))
annual_df %>% summarise(mean_tn_lbs = mean(tn_filt_lbs), sd_tn_lbs = sd(tn_filt_lbs), max_tn_lbs = max(tn_filt_lbs), min_tn_lbs = min(tn_filt_lbs))


```

### Figure 1: Map of Projects

```{r}


```

### Figure 2: Retention/release vs. volume vs. concentration

#### Volume vs. retention relationship

#### Figure
```{r}
# setup data
month_df_tp_filt <- month_df %>% filter(tp_filt_lbs >= 0) # separate out filtration (positive values)
month_df_tp_rel <- month_df %>% filter(tp_filt_lbs < 0) %>% mutate(tp_filt_lbs = abs(tp_filt_lbs)) # separate out release (negative values) and take absolute value
month_df_tn_filt <- month_df %>% filter(tn_filt_lbs >= 0) # separate out filtration (positive values)
month_df_tn_rel <- month_df %>%  filter(tn_filt_lbs < 0) %>% mutate(tn_filt_lbs = abs(tn_filt_lbs)) # separate out release (negative values) and take absolute value

# Make functions
log_minor_breaks <- function(lims) {
  min_exp <- floor(log10(lims[1]))
  max_exp <- ceiling(log10(lims[2]))
  unlist(lapply(min_exp:max_exp, \(e) (1:9) * 10^e))}

label_log_clean <- function(x) {
  sapply(x, function(val) {if (is.na(val))
      return("")
    exp <- log10(val)
    if (abs(exp - round(exp)) < 1e-10) {if (val < 1)
        return(format(val, scientific = FALSE, trim = TRUE))
      return(scales::comma(val, accuracy = 1))}  
    "" })}

# set up plot styles 

log_x_scale <- scale_x_log10(name = expression(bold("Inflow volume (m"^3*")")), limits = c(1, 1e6), 
  breaks = c(1, 10^(-10:10)), minor_breaks = log_minor_breaks, labels = label_log_clean)

log_y_scale <- scale_y_log10(limits = c(1e-4, 1e3), breaks = c(1e-6, 10^(-10:10)),
  minor_breaks = log_minor_breaks,labels = label_log_clean)

base_theme <- theme_bw() +
  theme(axis.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12))

cor_layer <- stat_cor(method = "pearson", aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),
  label.x.npc = 0, label.y.npc = 1, p.accuracy = 0.001)

# Make functions for creating plots
plot_volume <- function(data, yvar, ylab) {
  ggplot(data, aes(x = inflow_volume_m3, y = .data[[yvar]])) +
    geom_point() + geom_smooth(method = lm, color = "black") + log_x_scale +
    log_y_scale + labs(y = ylab) + annotation_logticks(sides = "bl") + cor_layer + base_theme}

plot_concentration <- function(data, xvar, yvar, xlab, ylab, smooth = FALSE) {
  p <- ggplot(data, aes( x = .data[[xvar]], y = .data[[yvar]])) +
    geom_point()
  if (smooth) {p <- p + geom_smooth(method = lm, color = "black")}
  p + log_y_scale + labs(x = xlab, y = ylab) + annotation_logticks(sides = "l") + cor_layer + theme_bw() +
    theme(axis.title = element_text(size = 14, face = "bold"), axis.text = element_text(size = 12))}

# Make Plots
p1 <- plot_volume(month_df_tp_filt, "tp_filt_lbs", "Total Phosphorus Filtration (lbs.)")
p2 <- plot_volume(month_df_tp_rel, "tp_filt_lbs", "Total Phosphorus Release (lbs.)")
p3 <- plot_concentration(month_df,"tp_in_mgL","tp_filt_lbs","Total Phosphorus concentration\n(mg/L) at Inflow", "Total Phosphorus Filtration (lbs.)")
p4 <- plot_volume(month_df_tn_filt, "tn_filt_lbs","Total Nitrogen Filtration (lbs.)")
p5 <- plot_volume(month_df_tn_rel,"tn_filt_lbs","Total Nitrogen Release (lbs.)")
p6 <- plot_concentration(month_df,"tn_in_mgL","tn_filt_lbs","Total Nitrogen concentration\n(mg/L) at Inflow", "Total Nitrogen Filtration (lbs.)", smooth = TRUE)

# Combine into final figure
fig_2 <- (p1 | p2 | p3) / (p4 | p5 | p6)

```
### Figure 3: Timing of retention/release
```{r}
# compile dfs and add seasons
dfs <- list(month_df_tp_rel  = month_df_tp_rel,
  month_df_tp_filt = month_df_tp_filt,
  month_df_tn_rel  = month_df_tn_rel,
  month_df_tn_filt = month_df_tn_filt)

dfs <- purrr::map(dfs, ~ .x %>%
  mutate(month_num = as.numeric(month), season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5 ~ "Spring",
      month %in% 6:8 ~ "Summer",
      month %in% 9:11 ~ "Fall"),
    season = factor(season,levels = c("Spring", "Summer", "Fall", "Winter"))) %>%
  arrange(Project, month_num))

# make sure that we're using the changed versions: 
month_df_tp_rel  <- dfs$month_df_tp_rel
month_df_tp_filt <- dfs$month_df_tp_filt
month_df_tn_rel  <- dfs$month_df_tn_rel
month_df_tn_filt <- dfs$month_df_tn_filt

# define season colors
season_colors <- c(
  Winter = "#A1CAF1",
  Spring = "#8DB600",
  Summer = "#F3C300",
  Fall   = "#882D17")

# define plots: 

make_plot <- function(data, yvar, title) {
  ggplot(data,aes(x = Project, y = .data[[yvar]], fill = season)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::label_percent()) +
    scale_fill_manual(values = season_colors, name = "Season") +
    labs(y = "Proportion", title = title, x = NULL) +
    theme_bw() + theme(legend.title = element_text(size = 12, face = "bold"),
      axis.title = element_text(size = 14, face = "bold"), axis.text = element_text(size = 12),
      axis.text.x = element_text(angle = 90)) }

# make plots
p1 <- make_plot(month_df_tp_filt, "tp_filt_lbs", "Proportion of Phosphorus Filtration")
p2 <- make_plot(month_df_tp_rel, "tp_filt_lbs","Proportion of Phosphorus Release")
p3 <- make_plot(month_df_tn_filt, "tn_filt_lbs", "Proportion of Nitrogen Filtration")
p4 <- make_plot(month_df_tn_rel, "tn_filt_lbs", "Proportion of Nitrogen Release")

# Combine together
fig_3 = (p1 / p2 / p3 / p4) + plot_layout(heights = 1, guides = "collect") & theme(legend.position = "right")
fig_3
```

### Figure 4: NEW ANALYSIS; Show when concentration does matter. Modelling??

Start with modelling this out. We're going to run a model for each project individually of retention ~ volume + concentration and see which is more important for each one. 

```{r}
month_df_tp_filt

# replace NAs with 0 when concentrations are not available: 
month_df_tp_filt <- month_df_tp_filt %>%  mutate(tp_in_mgL = coalesce(tp_in_mgL, 0))


project_model_stats <- month_df_tp_filt %>%
  group_by(Project) %>%
  group_modify(~{

    mod <- lm(
      tp_filt_lbs ~ inflow_volume_m3 + tp_in_mgL,
      data = .x
    )

    coefs <- summary(mod)$coefficients

    tibble(
      n = nrow(.x),
      r2 = summary(mod)$r.squared,
      inflow_volume_t = if ("inflow_volume_m3" %in% rownames(coefs))
        coefs["inflow_volume_m3", "t value"] else NA_real_,
      tp_in_t = if ("tp_in_mgL" %in% rownames(coefs))
        coefs["tp_in_mgL", "t value"] else NA_real_
    )
  }) %>%
  ungroup()

project_model_stats <- project_model_stats %>% mutate(diff_t = inflow_volume_t - abs(tp_in_t)) # take difference in t values (absolute value for concentration)


project_model_stats



```


### Figure 5: NEW ANALYSIS: What leads to nutrient release? 

```{r}
###### TP
# get the number of release events in the df for each month
month_df_tp_rel %>% dplyr::count(month, sort = FALSE)
# Make a monthly df with a column for whether there is TP retention (0) or release (1)
month_df_counts = month_df %>% mutate(ret_rel_tp = if_else(tp_filt_lbs < 0, 1, 0))
month_df_counts = month_df_counts %>% mutate(ret_rel_tp_text = if_else(tp_filt_lbs < 0, "release", "retain"))
# remove all values with 0 inflow volume
month_df_counts = month_df_counts %>% filter(tp_in_mgL != 0)
# make a model trying to predict TP retention or release
m1 <- glm(ret_rel_tp ~ inflow_volume_m3 + tp_in_mgL + month, data = month_df_counts, family = "binomial")
summary(m1)

###### TN 
# Make a monthly df with a column for whether there is TP retention (0) or release (1)
month_df_counts = month_df_counts %>% mutate(ret_rel_tn = if_else(tn_filt_lbs < 0, 1, 0))
# make a model trying to predict TN retention or release
m2 <- glm(ret_rel_tn ~ inflow_volume_m3 + tn_in_mgL + month, data = month_df_counts, family = "binomial")
summary(m2)
# release is more likely when there are low incoming concentrations

###### Plots
# make a db of predictions and then plot it using ggeffects
pred <- ggpredict(m1, terms = "tp_in_mgL")
p1 = ggplot(pred, aes(x = x, y = predicted)) + geom_line(linewidth = 1.2, color = "black") + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "TP In (mg/L)", y = "Predicted Probability of Release", title = "Predicted Probability of TP Release vs TP In") + theme_bw()
pred <- ggpredict(m2, terms = "tn_in_mgL")
p2 = ggplot(pred, aes(x = x, y = predicted)) + geom_line(linewidth = 1.2, color = "black") + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "TN In (mg/L)", y = "Predicted Probability of Release", title = "Predicted Probability of TN Release vs TN In") + theme_bw()
fig_5 = p1 + p2
fig_5
```

### Figure 6: Change over time (or lack thereof)
```{r}
# subset out only projects with 3+ years of data
long_term = annual_df %>%
  group_by(Project) %>%
  filter(n() >= 3)

# set color palette: 
focals_palette <-c("BROP" = '#0067A5', "BWLK" = '#F38400', "FORB" = '#008856',"LITP" = '#875692', "MAGM" = '#BE0032', "OAKE" = '#F6A600',"OAKW" = '#8DB600', "REDB" = '#B3446C', "SJRE" = '#654522', "SPRC" = '#2B3D26', "TIPC" = '#DCD300')

p1 = ggplot(long_term, aes(x = wateryear, y = tp_effic, color = Project)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3) +xlab("Water Year")+ ylab("Total Phosphorus Filtration Efficiency")+
   scale_color_manual(values = focals_palette) +
  theme_bw() +
  theme(legend.position = "top")+ theme(legend.title = element_text(size=12, face="bold"), axis.title = element_text(size=14, face="bold"), axis.text = element_text(size=12))

p2 = ggplot(long_term, aes(x = wateryear, y = tn_effic, color = Project)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3) +xlab("Water Year")+ ylab("Total Nitrogen Filtration Efficiency")+
   scale_color_manual(values = focals_palette) +
  theme_bw() +
  theme(legend.position = "top")+ theme(legend.title = element_text(size=12, face="bold"), axis.title = element_text(size=14, face="bold"), axis.text = element_text(size=12))

fig_6 = (p1 / p2) + plot_layout(heights = 1, guides = "collect") & theme(legend.position = "right")

```