---
title: "H2Ohio_Connectivity_Analysis"
author: "Kenny Anderson"
date: "2026-07-10"
output: html_document
---
```{r, setup, include=FALSE}
options(scipen =999)
```

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(ggpubr)
library(matrixStats)
library(scales)
library(patchwork)
library(scales)
library(Metrics)
library(rlang)
library(purrr)
library(stringr)
library(ggeffects)
library(ggspatial)  # for map annotations (north arrow, scale bar)
library(tidyterra)  # for handling and plotting raster data with ggplot2 (e.g., elevation)
library(pals)       # for color palettes (kelly)
library(sf)         # required to handle spatial data for maps
library(maps)       # for map data (e.g., county boundaries)
library(maptiles)   # for basemaps (e.g., Stamen Terrain)
library(ggrepel)    # for improved label placement in maps
library(magick)     # for image manipulation (e.g., adding logos to maps)
library(tigris)     # for Ohio boundary
```

## Load in Data

```{r}
annual_df = read.csv("Data/H2Ohio_2025_annualdf.csv")
month_df = read.csv("Data/H2Ohio_2025_monthdf.csv")

# calculate log values: 
month_df = month_df %>%
  mutate(loginflow = log10(inflow_volume_m3), logtpfilt = log10(tp_filt_lbs), logtpin = log10(tp_in_lbs), logtnfilt = log10(tn_filt_lbs), logdrpfilt = log10(drp_filt_lbs), lognoxfilt = log10(nox_filt_lbs), lognh3filt = log10(nh3_filt_lbs))

```


# Methods

### Wetland Summary Stats:
Get some summary stats on wetland hydrology
```{r}
loi_df = read.csv("Data/locations_of_interest_20260317.csv")
char_df = read.csv("Data/project character_20251119.csv")

# List all projects that have budgets
project_list <- unique(c(annual_df$Project, "MERW")) # add on MERW so we also capture values that don't list the three sub-projects individually
# subset out projects with budgets from char_df
char_df_budgets <- char_df %>%  filter(project_code %in% project_list) %>%  distinct(project_code, .keep_all = TRUE) # keep only the first row for each value

# We don't have EFLA and WALC in this db so we'll need to add values on manually. Also no data for UPPB so we'll need to add that too. 

# isolated pools present
char_df_budgets %>% summarise(n_yes = sum(isolated_pools == "Yes", na.rm = TRUE))
# number of lake connections
char_df_budgets %>% summarise(n_yes = sum(direct_coastal_connection == "Yes", na.rm = TRUE))
# number of stream connections
char_df_budgets <- char_df %>% filter(project_code %in% project_list) %>%  filter(str_detect(direct_riparian_connection, regex("Yes", ignore_case = TRUE)))
char_df_budgets %>% distinct(project_code) %>% pull(project_code)


# now do the same sorting with the LOI dataset to narrow in on the hydrology section
# subset out projects with budgets from loi_df that have direct tile inputs
loi_df_budgets <- loi_df %>%  filter(project_code %in% project_list) %>% filter(str_detect(hydrology, "direct tile input"))
# tell each project with direct tile inputs: 
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)
# subset out projects with budgets from loi_df that have floodplains/lotic influence
loi_df_budgets <- loi_df %>% filter(project_code %in% project_list) %>%  filter(str_detect(hydrology, regex("floodplain|lotic", ignore_case = TRUE)) | project_code == "EFLA") # manually include EFLA because it doesn't have data in the hydrology column
# tell each project with floodplains/lotic influence: 
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)


# subset out just pools
loi_df_budgets <- loi_df %>%  filter(project_code %in% project_list) %>% filter(str_detect(loi_cat_primary, regex("Pool|pool", ignore_case = TRUE)))
loi_df_budgets %>% distinct(project_code) %>% pull(project_code)
```

# Results

### Nutrient Budget Summary Stats: 

```{r}
# count the number of annual budgets that are positive and negative
annual_df %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
annual_df %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

annual_df %>% group_by(Project) %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
annual_df %>% group_by(Project) %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs <= 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

# count the number of months that are positive and negative
month_df %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
month_df %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

month_df %>% group_by(month) %>% summarise(n_positive = sum(tp_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tp_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tp_filt_lbs == 0, na.rm = TRUE))
month_df %>% group_by(month) %>% summarise(n_positive = sum(tn_filt_lbs > 0, na.rm = TRUE), n_negative = sum(tn_filt_lbs < 0, na.rm = TRUE), n_0 = sum(tn_filt_lbs == 0, na.rm = TRUE))

# mean and sd of annual retention: 
annual_df %>% summarise(mean_tp_lbs = mean(tp_filt_lbs), med_tp_lbs = median(tp_filt_lbs),sd_tp_lbs = sd(tp_filt_lbs), max_tp_lbs = max(tp_filt_lbs), min_tp_lbs = min(tp_filt_lbs))
annual_df %>% summarise(mean_tn_lbs = mean(tn_filt_lbs), med_tn_lbs = median(tn_filt_lbs), sd_tn_lbs = sd(tn_filt_lbs), max_tn_lbs = max(tn_filt_lbs), min_tn_lbs = min(tn_filt_lbs))

# mean and sd of nutrient uptake efficiency: 
annual_df %>% summarise(mean_tp_effic = mean(tp_effic,na.rm=T), med_tp_effic = median(tp_effic,na.rm=T), sd_tp_effic = sd(tp_effic), max_tp_effic = max(tp_effic), min_tp_effic = min(tp_effic))
annual_df %>% summarise(mean_tn_lbs = mean(tn_filt_lbs), sd_tn_lbs = sd(tn_filt_lbs), max_tn_lbs = max(tn_filt_lbs), min_tn_lbs = min(tn_filt_lbs))


```

### Figure 1: Map of Projects

```{r}
# Make a dataset of just the projects with budgets and the coordinates
loi_budgets_unique <- loi_df_budgets %>%
  distinct(project_code, .keep_all = TRUE) %>%
  mutate(loi_lon = sub("^--", "-", loi_lon)) %>% # take out weird double -
  select(project_code, loi_lat, loi_lon)

loi_budgets_unique$loi_lon = as.numeric(loi_budgets_unique$loi_lon)

# Download Ohio state boundary
ohio_sf <- states(cb = TRUE) |> 
  subset(NAME == "Ohio")

# If project does not have drone imagery, set a
# bounding box to be centered and add the padding
tile_bbox <- st_as_sfc(st_bbox(ohio_sf, crs = st_crs(4326)))

# Get tiles using maptiles library
# Zoom value has been adjusted from the original maps
# that used ggmap/get_map instead of maptiles library
tile <- get_tiles(tile_bbox, provider = "Esri.WorldGrayCanvas", zoom = 10)

# Fetch all US county spatial data
ohio_counties <- counties(state = "OH", cb = TRUE, class = "sf")

# (Optional but recommended) Transform to a projected coordinate system 
# that flattens the earth properly for Ohio (e.g., Ohio North/South 3734/3735)
ohio_counties_projected <- st_transform(ohio_counties, crs = 4326) 

# Build baseline plot
wy25_map <- 
  ggplot() +
    geom_spatraster_rgb(data = tile, alpha = 1) +
    geom_sf(data = ohio_sf, fill = NA, color = "black", linewidth = 0.5) +
    geom_sf(data = ohio_counties_projected, fill = NA, color = "gray60", linewidth = 0.5) +
    geom_point(data = loi_budgets_unique,
               aes(x = loi_lon, y = loi_lat), shape = 23,, fill = "gray30",
               stroke = 0.8, size = 3, show.legend = TRUE) +
    scale_fill_manual(name = "Monitored Projects\nin Water Year 2025") +
  geom_text_repel(data = loi_budgets_unique, aes(x = loi_lon, y = loi_lat, label = project_code), size = 3,
  box.padding = 0.3,point.padding = 0.2,max.overlaps = Inf)+
  theme_bw() + theme(axis.title = element_blank(), axis.text = element_blank(),legend.position = "top",
                     legend.title= element_text(size = 14, face = "bold"),
                     legend.text= element_text(size = 14)) + guides(fill = guide_legend(override.aes = list(size = 4))) +
    # Add scale bar to top-left corner
    annotation_scale(location = "bl", text_face = "bold", line_width = 2, text_cex = 1,
                     width_hint = 0.25, text_col = "black", line_col = "black",
                     pad_x = unit(0.015, "npc"), pad_y = unit(0.015, "npc"), style = "ticks") +
    # Add north arrow to top-righ corner
    annotation_north_arrow(
      location = "br", which_north = "true",
      pad_x = unit(0, "npc"), pad_y = unit(0.015, "npc"),
      style = north_arrow_fancy_orienteering(
        fill = c("black", "black"),
        line_col = NA,
        text_col = "black")) +
    coord_sf(expand = FALSE)
wy25_map  

```

### Figure 2: Retention/release vs. volume vs. concentration

#### Volume vs. retention relationship

#### Monthly Relationships
```{r}
# setup data
month_df_tp_filt <- month_df %>% filter(tp_filt_lbs >= 0) # separate out filtration (positive values)
month_df_tp_rel <- month_df %>% filter(tp_filt_lbs < 0) %>% mutate(tp_filt_lbs = abs(tp_filt_lbs)) # separate out release (negative values) and take absolute value
month_df_tn_filt <- month_df %>% filter(tn_filt_lbs >= 0) # separate out filtration (positive values)
month_df_tn_rel <- month_df %>%  filter(tn_filt_lbs < 0) %>% mutate(tn_filt_lbs = abs(tn_filt_lbs)) # separate out release (negative values) and take absolute value

# Make functions
log_minor_breaks <- function(lims) {
  min_exp <- floor(log10(lims[1]))
  max_exp <- ceiling(log10(lims[2]))
  unlist(lapply(min_exp:max_exp, \(e) (1:9) * 10^e))}

label_log_clean <- function(x) {
  sapply(x, function(val) {if (is.na(val))
      return("")
    exp <- log10(val)
    if (abs(exp - round(exp)) < 1e-10) {if (val < 1)
        return(format(val, scientific = FALSE, trim = TRUE))
      return(scales::comma(val, accuracy = 1))}  
    "" })}

# set up plot styles 

log_x_scale <- scale_x_log10(name = expression(bold("Inflow volume (m"^3*")")), limits = c(1, 1e6), 
  breaks = c(1, 10^(-10:10)), minor_breaks = log_minor_breaks, labels = label_log_clean)

log_y_scale <- scale_y_log10(limits = c(1e-4, 1e3), breaks = c(1e-6, 10^(-10:10)),
  minor_breaks = log_minor_breaks,labels = label_log_clean)

base_theme <- theme_bw() +
  theme(axis.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12))

cor_layer <- stat_cor(method = "pearson", aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),
  label.x.npc = 0, label.y.npc = 1, p.accuracy = 0.001)

# Make functions for creating plots
plot_volume <- function(data, yvar, ylab) {
  ggplot(data, aes(x = inflow_volume_m3, y = .data[[yvar]])) +
    geom_point() + geom_smooth(method = lm, color = "black") + log_x_scale +
    log_y_scale + labs(y = ylab) + annotation_logticks(sides = "bl") + cor_layer + base_theme}

plot_concentration <- function(data, xvar, yvar, xlab, ylab, smooth = FALSE) {
  p <- ggplot(data, aes( x = .data[[xvar]], y = .data[[yvar]])) +
    geom_point()
  if (smooth) {p <- p + geom_smooth(method = lm, color = "black")}
  p + log_y_scale + labs(x = xlab, y = ylab) + annotation_logticks(sides = "l") + cor_layer + theme_bw() +
    theme(axis.title = element_text(size = 14, face = "bold"), axis.text = element_text(size = 12))}

# Make Plots
p1 <- plot_volume(month_df_tp_filt, "tp_filt_lbs", "Total Phosphorus Filtration (lbs.)")
p2 <- plot_volume(month_df_tp_rel, "tp_filt_lbs", "Total Phosphorus Release (lbs.)")
p3 <- plot_concentration(month_df,"tp_in_mgL","tp_filt_lbs","Total Phosphorus concentration\n(mg/L) at Inflow", "Total Phosphorus Filtration (lbs.)")
p4 <- plot_volume(month_df_tn_filt, "tn_filt_lbs","Total Nitrogen Filtration (lbs.)")
p5 <- plot_volume(month_df_tn_rel,"tn_filt_lbs","Total Nitrogen Release (lbs.)")
p6 <- plot_concentration(month_df,"tn_in_mgL","tn_filt_lbs","Total Nitrogen concentration\n(mg/L) at Inflow", "Total Nitrogen Filtration (lbs.)", smooth = TRUE)

# Combine into final figure
fig_2 <- (p1 | p2 | p3) / (p4 | p5 | p6)
```

#### Leave one out analysis
```{r}
# setup dfs for dissolved nutrient filtration
month_df_drp_filt <- month_df %>% filter(drp_filt_lbs >= 0) # separate out filtration (positive values)
month_df_nox_filt <- month_df %>% filter(nox_filt_lbs >= 0) # separate out filtration (positive values)
month_df_nh3_filt <- month_df %>% filter(nh3_filt_lbs >= 0) # separate out filtration (positive values)

# make a function to run our LOO analysis for a specific variable: 
run_loo <- function(df, response_var, predictor_var = "loginflow") {

  # Replace -Inf with NA in the variables being modeled
  loo_df <- df %>%
    mutate(across(
      all_of(c(response_var, predictor_var)),
      ~ ifelse(is.infinite(.) & . < 0, NA, .)
    )) %>%
    na.omit()

  projects <- unique(loo_df$Project)

  loo.rmsesub <- rep(NA, length(projects))
  loo.rmseout <- rep(NA, length(projects))
  loo.adr2 <- rep(NA, length(projects))

  # Build model formula dynamically
  model_formula <- as.formula(
    paste(response_var, "~", predictor_var)
  )

  for(i in seq_along(projects)) {

    subdata <- loo_df[!(loo_df$Project %in% projects[i]), ]
    outdata <- loo_df[(loo_df$Project %in% projects[i]), ]

    loo_mod <- lm(model_formula, data = subdata)

    loo.rmsesub[i] <- rmse(
      subdata[[response_var]],
      predict(loo_mod, newdata = subdata)
    )

    loo.rmseout[i] <- rmse(
      outdata[[response_var]],
      predict(loo_mod, newdata = outdata)
    )

    loo.adr2[i] <- summary(loo_mod)$adj.r.squared

    print(
      paste(
        "Step", i,
        "Project:", projects[i],
        "RMSE M:", round(loo.rmsesub[i], 4),
        "RMSE R:", round(loo.rmseout[i], 4),
        "R2:", round(loo.adr2[i], 4)
      )
    )
  }

  results <- data.frame(
    project = projects,
    response = response_var,
    rmse_model = loo.rmsesub,
    rmse_removed = loo.rmseout,
    adj_r2 = loo.adr2
  )

  return(results)
}

# run all our variables
tp_results <- run_loo(month_df_tp_filt, "logtpfilt")
tn_results <- run_loo(month_df_tn_filt, "logtnfilt")
drp_results <- run_loo(month_df_drp_filt, "logdrpfilt")
nox_results <- run_loo(month_df_nox_filt, "lognoxfilt")
nh3_results <- run_loo(month_df_nh3_filt, "lognh3filt")

# un-log to get the RMSE for each variation for both the version with the full data included and removed
10^mean(tp_results$rmse_model)
10^mean(tp_results$rmse_removed)
10^mean(tn_results$rmse_model)
10^mean(tn_results$rmse_removed)
10^mean(drp_results$rmse_model)
10^mean(drp_results$rmse_removed)
10^mean(nox_results$rmse_model)
10^mean(nox_results$rmse_removed)
10^mean(nh3_results$rmse_model)
10^mean(nh3_results$rmse_removed)



loo_plots <- function(results, nutrient) {

  p_model <- ggplot(results, aes(x = project, y = rmse_model)) +
    geom_bar(stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90)) +
    ylab(paste("RMSE of model with site removed (log", nutrient, "Filtration (lbs))")) +
    xlab("Project Code")

  p_removed <- ggplot(results, aes(x = project, y = rmse_removed)) +
    geom_bar(stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90)) +
    ylab(paste("RMSE of model on removed site (log", nutrient, "Filtration (lbs))")) +
    xlab("Project Code")

  list(model = p_model, removed = p_removed)
}

tp_plots  <- loo_plots(tp_results,  "TP")
tn_plots  <- loo_plots(tn_results,  "TN")
drp_plots <- loo_plots(drp_results, "DRP")
nox_plots <- loo_plots(nox_results, "NOx")
nh3_plots <- loo_plots(nh3_results, "NH3")

figure_s1 <- ((tp_plots$model + tp_plots$removed) /
 (tn_plots$model + tn_plots$removed) /
 (nox_plots$model + nox_plots$removed) /
 (drp_plots$model + drp_plots$removed) /
 (nh3_plots$model + nh3_plots$removed)) +
  plot_annotation(title = "Leave-One-Site-Out RMSE Results", subtitle = "Model RMSE and Removed Site RMSE by Nutrient")
```

### Figure 3: Timing of retention/release
```{r}
# compile dfs and add seasons
dfs <- list(month_df_tp_rel  = month_df_tp_rel,
  month_df_tp_filt = month_df_tp_filt,
  month_df_tn_rel  = month_df_tn_rel,
  month_df_tn_filt = month_df_tn_filt)

dfs <- purrr::map(dfs, ~ .x %>%
  mutate(month_num = as.numeric(month), season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5 ~ "Spring",
      month %in% 6:8 ~ "Summer",
      month %in% 9:11 ~ "Fall"),
    season = factor(season,levels = c("Spring", "Summer", "Fall", "Winter"))) %>%
  arrange(Project, month_num))

# make sure that we're using the changed versions: 
month_df_tp_rel  <- dfs$month_df_tp_rel
month_df_tp_filt <- dfs$month_df_tp_filt
month_df_tn_rel  <- dfs$month_df_tn_rel
month_df_tn_filt <- dfs$month_df_tn_filt

# define season colors
season_colors <- c(
  Winter = "#A1CAF1",
  Spring = "#8DB600",
  Summer = "#F3C300",
  Fall   = "#882D17")

# define plots: 

make_plot <- function(data, yvar, title) {
  ggplot(data,aes(x = Project, y = .data[[yvar]], fill = season)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::label_percent()) +
    scale_fill_manual(values = season_colors, name = "Season") +
    labs(y = "Proportion", title = title, x = NULL) +
    theme_bw() + theme(legend.title = element_text(size = 12, face = "bold"),
      axis.title = element_text(size = 14, face = "bold"), axis.text = element_text(size = 12),
      axis.text.x = element_text(angle = 90)) }

# make plots
p1 <- make_plot(month_df_tp_filt, "tp_filt_lbs", "Proportion of Phosphorus Filtration")
p2 <- make_plot(month_df_tp_rel, "tp_filt_lbs","Proportion of Phosphorus Release")
p3 <- make_plot(month_df_tn_filt, "tn_filt_lbs", "Proportion of Nitrogen Filtration")
p4 <- make_plot(month_df_tn_rel, "tn_filt_lbs", "Proportion of Nitrogen Release")

# Combine together
fig_3 = (p1 / p2 / p3 / p4) + plot_layout(heights = 1, guides = "collect") & theme(legend.position = "right")
fig_3
```

### Figure 4: NEW ANALYSIS; Show when concentration does matter. Modelling??

Start with modelling this out. We're going to run a model for each project individually of retention ~ volume + concentration and see which is more important for each one. 

```{r}
# run a model for each individual site and take the r2, and t values for concentration and volume and put in a df
get_project_model_stats <- function(df, response_var, concentration_var,volume_var = "inflow_volume_m3") {
  # Replace NAs in concentration variable with 0
  df <- df %>% mutate(!!concentration_var := coalesce(.data[[concentration_var]], 0))

  df %>%
    group_by(Project) %>%
    group_modify(~{ form <- as.formula(paste(response_var, "~", volume_var, "+", concentration_var))

      mod <- lm(form, data = .x)
      coefs <- summary(mod)$coefficients

      tibble(n = nrow(.x), r2 = summary(mod)$r.squared,
        inflow_volume_t = if (volume_var %in% rownames(coefs))
            coefs[volume_var, "t value"]
          else NA_real_, concentration_t = if (concentration_var %in% rownames(coefs))
            coefs[concentration_var, "t value"]
          else NA_real_ )}) %>%  ungroup()
}

# run for all the different nutrients
tp_stats <- get_project_model_stats(month_df_tp_filt, response_var = "tp_filt_lbs", concentration_var = "tp_in_mgL")
tn_stats <- get_project_model_stats(month_df_tn_filt, response_var = "tn_filt_lbs", concentration_var = "tn_in_mgL")
drp_stats <- get_project_model_stats(month_df_drp_filt, response_var = "drp_filt_lbs", concentration_var = "drp_in_mgL")
nox_stats <- get_project_model_stats(month_df_nox_filt, response_var = "nox_filt_lbs", concentration_var = "nox_in_mgL")
nh3_stats <- get_project_model_stats(month_df_nh3_filt, response_var = "nh3_filt_lbs", concentration_var = "nh3_in_mgL")

```
#### Heatmap figure
```{r}
# remove models that don't fit right because there's not enough data
tp_stats = tp_stats %>% mutate(inflow_volume_t = if_else(Project == "MERW1", NA, inflow_volume_t))
tn_stats = tn_stats %>% mutate(inflow_volume_t = if_else(Project == "MERW1", NA, inflow_volume_t))
nh3_stats = nh3_stats %>% mutate(inflow_volume_t = if_else(Project %in% c("MERW1", "MERW2","MONW"), NA, inflow_volume_t))
nox_stats = nox_stats %>% mutate(inflow_volume_t = if_else(Project %in% c("MERW1", "MERW2"), NA, inflow_volume_t))


# pivot the df longer to setup for plotting
tp_stats_long <- tp_stats %>%  pivot_longer(cols = ends_with("_t"), names_to = "variable", names_pattern = "(.*)_t", values_to = "t_value")
tn_stats_long <- tn_stats %>%  pivot_longer(cols = ends_with("_t"), names_to = "variable", names_pattern = "(.*)_t", values_to = "t_value")
drp_stats_long <- drp_stats %>%  pivot_longer(cols = ends_with("_t"), names_to = "variable", names_pattern = "(.*)_t", values_to = "t_value")
nox_stats_long <- nox_stats %>%  pivot_longer(cols = ends_with("_t"), names_to = "variable", names_pattern = "(.*)_t", values_to = "t_value")
nh3_stats_long <- nh3_stats %>%  pivot_longer(cols = ends_with("_t"), names_to = "variable", names_pattern = "(.*)_t", values_to = "t_value")

# calculate absolute value because negative t values are just as meaningful as positive
tp_stats_long$abs_t_value = abs(tp_stats_long$t_value)
tn_stats_long$abs_t_value = abs(tn_stats_long$t_value)
drp_stats_long$abs_t_value = abs(drp_stats_long$t_value)
nox_stats_long$abs_t_value = abs(nox_stats_long$t_value)
nh3_stats_long$abs_t_value = abs(nh3_stats_long$t_value)



# make the plot
p1 <- ggplot(tp_stats_long, aes(Project, variable, fill = abs_t_value)) +
  geom_tile() + coord_equal() + scale_fill_gradientn(colours = c("white", "gray40", "black"), trans = "log10",
    limits = c(1, 150), oob = scales::squish, breaks = c(1, 10, 150), name = "t value") +  theme_bw()+ylab("")+xlab("")+ggtitle("Total Phosphorus")
p2 <- ggplot(tn_stats_long, aes(Project, variable, fill = abs_t_value)) +
  geom_tile() + coord_equal() + scale_fill_gradientn(colours = c("white", "gray40", "black"), trans = "log10",
    limits = c(1, 150), oob = scales::squish, breaks = c(1, 10, 150), name = "t value") +  theme_bw()+ylab("")+xlab("")+ggtitle("Total Nitrogen")
p3 <- ggplot(drp_stats_long, aes(Project, variable, fill = abs_t_value)) +
  geom_tile() + coord_equal() + scale_fill_gradientn(colours = c("white", "gray40", "black"), trans = "log10",
    limits = c(1, 150), oob = scales::squish, breaks = c(1, 10, 150), name = "t value") +  theme_bw()+ylab("")+xlab("")+ggtitle("Dissolved Reactive Phosphorus")
p4 <- ggplot(nox_stats_long, aes(Project, variable, fill = abs_t_value)) +
  geom_tile() + coord_equal() + scale_fill_gradientn(colours = c("white", "gray40", "black"), trans = "log10",
    limits = c(1, 150), oob = scales::squish, breaks = c(1, 10, 150), name = "t value") +  theme_bw()+ylab("")+xlab("")+ggtitle("NoX")
p5 <- ggplot(nh3_stats_long, aes(Project, variable, fill = abs_t_value)) +
  geom_tile() + coord_equal() + scale_fill_gradientn(colours = c("white", "gray40", "black"), trans = "log10",
    limits = c(1, 150), oob = scales::squish, breaks = c(1, 10, 150), name = "t value") +  theme_bw()+ylab("")+xlab("")+ggtitle("Ammonium")

fig_4 = (p1 / p2 / p3 / p4/ p5) + plot_layout(heights = 1, guides = "collect") & theme(legend.position = "right")
fig_4

```

### Figure 5: NEW ANALYSIS: What leads to nutrient release? 

```{r}
###### TP
# get the number of release events in the df for each month
month_df_tp_rel %>% dplyr::count(month, sort = FALSE)
# Make a monthly df with a column for whether there is TP retention (0) or release (1)
month_df_counts = month_df %>% mutate(ret_rel_tp = if_else(tp_filt_lbs < 0, 1, 0))
month_df_counts = month_df_counts %>% mutate(ret_rel_tp_text = if_else(tp_filt_lbs < 0, "release", "retain"))
# remove all values with 0 inflow volume
month_df_counts = month_df_counts %>% filter(tp_in_mgL != 0)
# make a model trying to predict TP retention or release
m1 <- glm(ret_rel_tp ~ inflow_volume_m3 + tp_in_mgL + month, data = month_df_counts, family = "binomial")
summary(m1)

###### TN 
# Make a monthly df with a column for whether there is TP retention (0) or release (1)
month_df_counts = month_df_counts %>% mutate(ret_rel_tn = if_else(tn_filt_lbs < 0, 1, 0))
# make a model trying to predict TN retention or release
m2 <- glm(ret_rel_tn ~ inflow_volume_m3 + tn_in_mgL + month, data = month_df_counts, family = "binomial")
summary(m2)
# release is more likely when there are low incoming concentrations

###### Plots
# make a db of predictions and then plot it using ggeffects
pred <- ggpredict(m1, terms = "tp_in_mgL")
p1 = ggplot(pred, aes(x = x, y = predicted)) + geom_line(linewidth = 1.2, color = "black") + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "TP In (mg/L)", y = "Predicted Probability of Release", title = "Predicted Probability of TP Release vs TP In") + theme_bw()
pred <- ggpredict(m2, terms = "tn_in_mgL")
p2 = ggplot(pred, aes(x = x, y = predicted)) + geom_line(linewidth = 1.2, color = "black") + geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "TN In (mg/L)", y = "Predicted Probability of Release", title = "Predicted Probability of TN Release vs TN In") + theme_bw()
fig_5 = p1 + p2
fig_5
```

### Figure 6: Change over time (or lack thereof)
```{r}
# subset out only projects with 3+ years of data
long_term = annual_df %>%
  group_by(Project) %>%
  filter(n() >= 3)

# set color palette: 
focals_palette <-c("BROP" = '#0067A5', "BWLK" = '#F38400', "FORB" = '#008856',"LITP" = '#875692', "MAGM" = '#BE0032', "OAKE" = '#F6A600',"OAKW" = '#8DB600', "REDB" = '#B3446C', "SJRE" = '#654522', "SPRC" = '#2B3D26', "TIPC" = '#DCD300')

p1 = ggplot(long_term, aes(x = wateryear, y = tp_effic, color = Project)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3) +xlab("Water Year")+ ylab("Total Phosphorus Filtration Efficiency")+
   scale_color_manual(values = focals_palette) +
  theme_bw() +
  theme(legend.position = "top")+ theme(legend.title = element_text(size=12, face="bold"), axis.title = element_text(size=14, face="bold"), axis.text = element_text(size=12))

p2 = ggplot(long_term, aes(x = wateryear, y = tn_effic, color = Project)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3) +xlab("Water Year")+ ylab("Total Nitrogen Filtration Efficiency")+
   scale_color_manual(values = focals_palette) +
  theme_bw() +
  theme(legend.position = "top")+ theme(legend.title = element_text(size=12, face="bold"), axis.title = element_text(size=14, face="bold"), axis.text = element_text(size=12))

fig_6 = (p1 / p2) + plot_layout(heights = 1, guides = "collect") & theme(legend.position = "right")

```

### Table 4: Annual retention model

```{r}
# add in mean SPSC

# add in veg data where available

# 


```

